Galois Slicing as Automatic Differentiation

### 2025-07-11 main/models.tex
[ tex ] some final edits

The category $\MeetSLat$ satisfies the hypoetheses of
\thmref{fam-closed}, so:
\begin{corollary}
  $\Fam(\MeetSLat)$ is cartesian closed and has all coproducts.
\end{corollary}
Unfortunately, neither of the $\LatGal$ or $\FinVect$ satisfy the
hypotheses of this theorem, because neither of them have infinite
products. We will consider ways to rectify this below in
\secref{fixing-completeness}.

\begin{remark}
  \label{rem:hermida-exponentials}
  There is another construction of internal homs on $\Fam(\cat{C})$
  arising from the use of fibrations for categorical logical
  relations, due to \citet[Corollary 4.12]{hermida99}. If we assume
  that $\cat{C}$ is itself cartesian closed and has all products, then
  we could construct an internal hom as:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (X \to Y, \lambda f. \Pi_{x : X}.\,\partial X(x) \to \partial Y(f\,x))
  \end{displaymath}
  However, for the purposes of modelling differentiable programs, this
  is fatally flawed in that neither of $\LatGal$ nor $\FinVect$ are
  cartesian closed, and there is no way of making them so without
  losing the property of being able to conjunct or add tangents, as we
  shall see below. We will implicitly use Hermida's construction in
  our definability proof in \secref{definability}, where we use a
  logical relations argument to show that every morphism definable in
  the higher order language is also first-order definable.
\end{remark}

@@

\begin{example}
  The following are $\CMon$-enriched and have finite products, and
  hence biproducts:
  \begin{enumerate}[leftmargin=\enummargin]
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set and the one-element lattice for the
    terminal/initial/zero object.
  \item $\MeetSLat$ and $\JoinSLat$ are both $\CMon$-enriched and have
    finite products for similar reasons to $\LatGal$.
  \end{enumerate}
\end{example}

@@

In the case of $\FinVect$, the solution is to expand to the category
of all vector spaces $\Vect$, where infinite direct products
exist. Note that these infinite products are not biproducts because
the vector space operations themselves are finitary. This is the
solution that \cite{vakar22} uses for the semantics of forward
($\Fam(\Vect)$) and reverse ($\Fam(\Vect^\op)$) automatic
differentiation for higher order programs. Since the forward and
reverse derivatives of a smooth map are intrinsically defined,
\citet{vakar22}'s correctness theorem shows that, for programs with
first-order type, the interpretation in $\Fam(\Vect)$ correctly yields
the forward derivative of the defined function on the reals (and
reverse derivative for $\Fam(\Vect^\op)$).

@@

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$\mathbb{R}_{\mathit{intv}} = (\mathbb{R}, \lambda x.\,\{[l,u] \mid l
\leq x \leq u\} \cup \{\bot\})$ where the lattices of intervals are
reverse ordered by inclusion with $\bot$ at the bottom. Then,
following the examples in \exref{intervals-and-maxima-elements} and
\exref{stable-functions}, we can define addition, negation, and
scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,{-}x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
(we only define the forward and backward maps on intervals, their
behaviour on $\bot$ is determined.)  Scaling by negative numbers is
also possible with swapping of bounds, as is multiplication.

@@

We have seen that the category $\Fam(\MeetSLat \times \JoinSLat^\op)$,
which is a cartesian closed category with all coproducts, is enough to
interpret the total higher-order language we define in the next
section with primitive types and operations defined in
$\Fam(\LatGal)$. However, we are not guaranteed by construction that
at first-order type, the interpretations are in fact Galois
connections. We will rectify this in \secref{definability} using a
logical relations construction.

>>>>

The category $\MeetSLat$ satisfies the hypotheses of
\thmref{fam-closed}, so:
\begin{corollary}
  $\Fam(\MeetSLat)$ is cartesian closed and has all coproducts.
\end{corollary}
Unfortunately, neither $\LatGal$ nor $\FinVect$ satisfy the
hypotheses of this theorem, because neither of them have infinite
products. We will consider ways to rectify this below in
\secref{fixing-completeness}.

\begin{remark}
  \label{rem:hermida-exponentials}
  There is another construction of internal homs on $\Fam(\cat{C})$
  arising from the use of fibrations for categorical logical
  relations, due to \citet[Corollary 4.12]{hermida99}. If we assume
  that $\cat{C}$ is itself cartesian closed and has all products, then
  we could construct an internal hom as:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (X \to Y, \lambda f. \Pi_{x : X}.\,\partial X(x) \to \partial Y(f\,x))
  \end{displaymath}
  However, for the purposes of modelling differentiable programs, this
  is fatally flawed in that neither $\LatGal$ nor $\FinVect$ are
  cartesian closed, and there is no way of making them so without
  losing the property of being able to conjunct or add tangents, as we
  shall see below. We will implicitly use Hermida's construction in
  our definability proof in \secref{definability}, where we use a
  logical relations argument to show that every morphism definable in
  the higher order language is also first-order definable.
\end{remark}

@@

\begin{example}
  The following are $\CMon$-enriched and have finite products, and
  hence biproducts:
  \begin{enumerate}[leftmargin=\enummargin]
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set and the one-element lattice for the
    terminal/initial/zero object.
  \item $\MeetSLat$ and $\JoinSLat$ are both $\CMon$-enriched and have
    finite products similar to $\LatGal$.
  \end{enumerate}
\end{example}

@@

In the case of $\FinVect$, the solution is to expand to the category
of all vector spaces $\Vect$, where infinite direct products
exist. Note that these infinite products are not biproducts because
the vector space operations themselves are finitary. This is the
solution that \citet{vakar22} use for the semantics of forward
($\Fam(\Vect)$) and reverse ($\Fam(\Vect^\op)$) automatic
differentiation for higher order programs. Since the forward and
reverse derivatives of a smooth map are intrinsically defined,
\citet{vakar22}'s correctness theorem shows that, for programs with
first-order type, the interpretation in $\Fam(\Vect)$ correctly yields
the forward derivative of the defined function on the reals (and
reverse derivative for $\Fam(\Vect^\op)$).

@@

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$\mathbb{R}_{\mathit{intv}} = (\mathbb{R}, \lambda x.\,\{[l,u] \mid l
\leq x \leq u\} \cup \{\bot\})$ where the lattices of intervals are
reverse ordered by inclusion with $\bot$ at the bottom. Then,
following the examples in \exref{intervals-and-maxima-elements} and
\exref{stable-functions}, we can define addition, negation, and
scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,{-}x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
(we only define the forward and backward maps on intervals, their
behaviour on $\bot$ is determined.)  Scaling by negative numbers is
also possible with swapping of bounds, as is multiplication. We will
see an example of the use of these operations in
\secref{language:examples}.

@@

We have seen that the category $\Fam(\LatGal)$ has enough structure to
express useful approximation maps at first-order and that
$\Fam(\MeetSLat \times \JoinSLat^\op)$, which is a cartesian closed
category with all coproducts, is enough to interpret the total
higher-order language we define in the next section with primitive
types and operations defined in $\Fam(\LatGal)$. However, we are not
guaranteed by construction that at first-order type, the
interpretations are in fact Galois connections. We will rectify this
in \secref{definability} using a logical relations construction.
